A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-24 text/discussion.Rmd
update discussion of histological subtypes, metastasis, add machine learning references

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomic, proteomic data and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways represented by our classes (**figure 1a-k**). Each
of these classes may represent unique combinations of pathway disruptions with
increased fitness regardless of tissue of origin, which then converge on end
stage cancer by accumulating complementary mutations to the other classes
(**figure 1l** and **Table 2**).

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1h** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
see **supplementary figure S1**) are unevenly distributed, though we could not
identify cancer type that was exclusive to a single tumor class. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental
Figure S2**). Indeed, it would be surprising if we did not observe bias for some
cancers, as it has long been known that certain treatment regimens are likely to
work for a significant fraction of patients of a given cancer type. Nonetheless,
the basis for some tumors being treatment-refractory in spite of clinical
indications remains elusive. Our analysis suggests the possibility of
identifying more informative molecular, histological or cellular subtypes and
could form a basis for future separation of patients into different precision
treatment regimens.

It could be that the classes we identified are consistent with a signature of
overall disease progression, such that successive, adjacent classes on the UMAP
projection exhibit increasing specificity. We identified no such trend (see
**figure 1k**) and each class instead was associated with unique combinations of
pathways (as in **figure 3**). In our view, these observations suggest that
cancers have available to them a limited number of stable configurations with
respect to pathway disruption which are consistent with increased fitness and
therefore positive selection.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 8, which for which we did not find distinctive associations with
pathways. Class 1 has a relatively low level of mutated pathways overall,
although it is broadly enriched in many of the same tumor-promoting pathways
common to the other groups. This likely does not reflect a true difference in
stages of tumor evolution, as our data show clearly that this class is as likely
to contain stage IV metastatic tumors as it is to contain those of stage I. We
speculate this class represents a group of tumors with large aberrations in
methylation, copy number or other genome rearrangements which we excluded from
the present analysis. In the future, we hope to incorporate these other data
into a pathway-centric analysis as we have done here for point mutations and
indels. Now that we have established this as a feasible and potentially fruitful
approach to understanding the cancer genome, it may be worth expanded effort in
this area. In addition, Class 8 represents a group of hyper-mutated tumors that
harbor so many mutations that virtually no pathway is unaffected. It seems
likely that a significant fraction of the samples flagged as “mutant” for each
pathway are actually unaffected, consistent with those samples have low
information content. Similar to class 1 we speculate that a future analysis of
other data types, including improved filtering of mutations across all classes,
may resolve these tumors into improved classifications.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 5**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class _and_
cancer type fit better than models in which these variables are independent
factors. However, considering that within each cancer type there are different
clinical standards of care, and even within classes of drugs the preferred
treatment can very between cancers, it makes sense in hindsight that they should
all be different. Contrast the situation with ovarian vs breast cancer, which
are both hormonally driven cancers, for example. Ovarian cancer has but one main
treatment, platinum, whereas breast cancer patients have a variety of treatment
regimens based on molecular subtype and other factors. It is easy to see how
standard of care for one cancer and class could produce different outcomes than
the same class of a different cancer type under the circumstances.
Unfortunately, given the diversity drug classes and treatments (including
non-chemical treatments such as surgery and palliative care), we lack sufficient
power in the current study to explore these variables in the GDC data. It is our
hope that more focused future studies will help to distinguish between
treatment-specific effects or survival given different pathway disruption
classes.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (see **figures 2** and
**supplementary figure S1**) and often are the result of aberrations involving
common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
Class-specific evolution must involve the acquisition of combinatorial
disruptions to pathways that may individually be harmful but together produce
more specialized cancer phenotypes and increased fitness (as in **figure 4** and
**tables 1** and **ST2**). Importantly, our work does not distinguish between
the order in which these mutations should accumulate. However, since many of the
genes in the non-specific pathways involve genome stability, it is reasonable to
hypothesize that such mutations promote the acquisition of class specific
defects, and natural selection produces the classes we observed.

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity, both in terms of evolving subclones with different or
passively diverging mutation profiles which then serve as a substrate for futher
selection [@PCAWG-MAIN; @PCAWG-EVO]. Moreover consortium samples are likely to
contain contamination from support tissue and inflammatory and immune cells of
the innate and adaptive immune systems, potentially harboring cancer supporting
mutations [@bhowmick_review]. We think it will be instructive to explore these
ideas in the context of single cell experiments.

@@

We report that pathway enrichment across all classes (pooled) yielded generally
lower effect sizes and larger p-values than the cluster specific analysis,
suggesting that signal is diluted when the samples are analyzed this way, and
supporting the view that metastasis has class-specific requirements. It should
be noted, however, that in spite of this lack of global enrichment, some of the
cluster specific pathways reflect known metastatic associations, _e.g._ pathways
associated with asparagine and aspartate metabolism in class 10
[@knott_asparagus]. Since the number of metastatic tumors is relatively low this
part of our analysis is likely underpowered and subject to expanded analysis
with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, since (again) we don’t
know the order in which they were acquired in individual tumor samples. The fact
that most metastatic enrichment is cluster-specific and has a tendency to
overlap with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. This is consistent with the observation
that tumors increase diversity of driver mutations in later stages [@PCAWG-EVO].
When we discover the requirements of each cluster with respect to metastasis we
can target them therapeutically and prevent further adaptation.

>>>>

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of TCGA consortium
papers present a view largely, and surprisingly, consistent with this hypothesis
[@hoadleyMultiplatformAnalysis2014; @hoadleyCellofOriginPatternsDominate2018].
Perhaps owing to the intractable complexity of genomic, proteomic data and
patient metadata in all its forms, the inescapable conclusion thus far is that
tissue-of-origin remains the most important driver of tumor characteristics at
every scale and by every measure. Our observations contrast with this view, and
instead support an interpretation of publicly available data in which all tumors
manifest one of a limited number of metastable phenotypes resulting from
disruptions in basic pathways represented by our classes (**figure 1**). Each
of these classes may represent unique combinations of pathway disruptions with
increased fitness regardless of tissue of origin, which then converge on end
stage cancer by accumulating complementary mutations to the other classes
(**figures 1L and M** and **Table 2**).

We find that most tumors segregate into one of twelve biologically distinct
clusters, regardless of the cancer type (**figure 1G,I** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 8 and 9, and Kidney Chromophobe in cluster 1, see **supplementary
figure S1**) are unevenly distributed, though we could not identify cancer type
that was exclusive to a single tumor class. Only cancer types with the fewest
samples were found to be absent from one or more classes at all. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1J** and **Supplemental
Figure S2**). This result implies that many, if not most, histological subtypes
primarily reflect differences in cell-of-origin, rather than fundamental
differences in proximal cause. It has been hypothesized that differences in the
molecular regulators of development in the array of cell types present in the
breast epithelium drive the various histological phenotypes [@breast-subtypes].
Consistent with this view we found that breast tumor samples of the Luminal A
subtype were heavily biased toward membership in clusters 1 and 7, and basal
tumor samples were biased toward cluster 6 (**figure 1J**). However, both
subtypes contained samples in every other class without exception, and Luminal B
and Her2 positive samples are distributed across clusters. We interpret these
data to mean that cell-of-origin molecular machinery can predispose certain cell
types within the breast epithelium to forming tumors of one class or another but
is not determinative; this view is compatible with the previously stated
hypothesis but opens the way for a more granular view of individual tumors.
Indeed, it would be surprising if we did not observe bias for some cancers and
subtypes amongst our classes, since some treatment regimens have greater
efficacy for a significant fraction of patients of a given cancer/histological
subtype. Nonetheless, the basis for some tumors being treatment-refractory in
spite of receiving the standard of clinical care for diagnostic markers remains
elusive. Doubtless some of this is due to bad luck, as tumors can metastasis and
remain dormant years before they are detected at distal sites [@hallmarks_II], or
resistant clones may have already arisen at undetectable levels, but our
analysis suggests the possibility of identifying more informative molecular,
histological or cellular subtypes that could form a basis for future separation
of patients into different precision treatment regimens.

It could be that the classes we identified are consistent with a signature of
overall disease progression, such that successive, adjacent classes on the UMAP
projection exhibit increasing specificity. We identified no such trend (see
**figure 1H**) and each class instead was associated with unique combinations of
pathways (as in **figures 2B, C**). In our view, these observations suggest that
cancers have available to them a limited number of stable configurations with
respect to pathway disruption which are consistent with increased fitness and
therefore positive selection.

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 10, which for which we did not find distinctive associations with
pathways (**figure 2B, C**). Class 1 has a relatively low proportion of mutated
pathways across the board, although it is broadly enriched in many of the same
tumor-promoting pathways common to the other groups. This likely does not
reflect a true difference in stages of tumor evolution, as our data show clearly
that this class is as likely to contain stage IV metastatic tumors as it is to
contain those of stage I (**figure 1L, M**). This class likely represents a
group of tumors with large aberrations in methylation, copy number alterations,
structural variants, or combinations thereof which we excluded from our
analysis. Consistent with this, kidney chromophobe and thyroid cancers have high
proportions of SVs _vs._ other variant types [@PCAWG-MAIN] and are heavily
skewed to cluster 1 membership (**Supplementary figure S1**). Likewise, Class 8
represents a group of hyper-mutated tumors that harbor so many mutations that
virtually no pathway is unaffected. It seems likely that a significant fraction
of the samples flagged as “mutant” for each pathway are actually unaffected,
consistent with those samples have low information content. This could be
addressed with application of ever-increasingly sophisticated bioinformatics
algorithms for filtering likely passenger mutations [ _e.g._ @spliceAI;
@deep-learning-clinical-variants]. In the future, we hope to incorporate these
other data into a pathway-centric analysis as we have done here for point
mutations and indels.

@@

By modeling CDC longevity data as a baseline risk function we showed that each
of our pathway disruption classes exhibit cancer-type specific effects on
expected survival (**figure 4**). Contrary to what one might expect at the
outset, we found that models assuming different survival based on class _and_
cancer type fit better than models in which these variables are independent
factors. However, considering that within each cancer type there are different
clinical standards of care, and even within classes of drugs the preferred
treatment can very between cancers, it makes sense in hindsight that they should
all be different. Contrast the situation with ovarian vs breast cancer, which
are both hormonally driven cancers, for example. Ovarian cancer has but one main
treatment, platinum, whereas breast cancer patients have a variety of treatment
regimens based on molecular subtype and other factors. It is easy to see how
standard of care for one cancer and class could produce different outcomes than
the same class of a different cancer type under the circumstances.
Unfortunately, given the diversity drug classes and treatments (including
non-chemical treatments such as surgery and palliative care), we lack sufficient
power in the current study to explore these variables in the TCGA data. It is our
hope that more focused future studies will help to distinguish between
treatment-specific effects or survival given different pathway disruption
classes.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (**supplementary figure S1**)
and often are the result of aberrations involving common driver genes such as
BRCA1/2, MMR genes, mitotic checkpoints, Cohesion complexes, _etc._
(**supplementary figure S3** and **supplemental file 1**). Class-specific
evolution must involve the acquisition of combinatorial disruptions to pathways
that may individually be harmful but together produce more specialized cancer
phenotypes and increased fitness (as in **figure 4** and **tables 1** and
**ST2**). Importantly, our work does not distinguish between the order in which
these mutations should accumulate. This could be addressed in a future study by
integrating our analysis with evolutionary analysis of clonality, drawing
inferences from variant allele frequencies as in @PCAWG-EVO. However, since many
of the genes in the non-cluster-specific pathways involve genome stability, it
is reasonable to hypothesize that such mutations promote the acquisition of
class-specific defects via random mutation and natural selection, thus producing
the classes we observed. Indeed, a subset of known oncogenic driver mutations
are highly enriched in early arising clones, whereas later arising clones have
much greater diversity in driver mutations [@PCAWG-EVO]. Moreover, driver genes
that are known to be responsible for discrete mutation signatures such as
APOBEC, BRCA1 and BRCA2 produce mutational hotspots reflecting the varying
selective pressures in different tissues [@PCAWG-MAIN].

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity, thus consortium samples are likely to
contain contamination from support tissue and inflammatory and immune cells of
the innate and adaptive immune systems, potentially harboring cancer supporting
mutations [@bhowmick_review]. We think it will be instructive to explore these
ideas in the context of single cell experiments.

@@

We report that pathway enrichment in metastatic tumors across all classes
(pooled) yielded generally lower effect sizes and larger p-values than the
cluster specific analysis, suggesting that signal is diluted when the samples
are analyzed this way, and supporting the view that metastasis has
class-specific requirements. Since the number of metastatic tumors is relatively
low this part of our analysis is likely underpowered and subject to expanded
analysis with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, since we don’t know the
order in which they were acquired in individual tumor samples. The fact that
most metastatic enrichment is cluster-specific and has a tendency to overlap
with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. Some clusters were characterized by
cluster-specific pathways that seem superficially unrelated; for example cluster
3 has mutations impacting protein translation and turnover, whereas cluster 5
has defects in RNA metabolism and clusters 7, 8 and 9 have defects in the more
traditional extra-cellular signaling pathways (as exemplified in **table 1**).
It may be that the primary effects of mutations in polymer metabolism in cluster
3 and 5 point to currently unappreciated mechanisms impacting hallmark pathways.
Clusters 3 and 5 are also marked by a decidedly different set of pathways in
metastasis. Clusters 3 and 5 notably have (in addition to EGFR signaling)
mutations in glucagon and glycogen metabolism, respectively, in their metastases
(**table 2**), consistent with energy metabolism as one of the new hallmarks
[@hallmarks_II], but also distinguishing these classes from the others in the
order or nature of hallmark acquisition. We submit that when we discover the
requirements of each cluster with respect to metastasis we may be able to target
them therapeutically and prevent further adaptation.
